Wormtable: A data structure for genome scale data

### 2013-07-04 docs/paper/paper.tex

% Databases solve these problems
The classical approach to working with large volumes of 
structured data is to use a database, which solve these 
problems very effectively.
Data is typically encoded in an efficient binary 
format, saving substantial space and giving better precision when 
dealing with floating point formats. Records are stored in a
B-tree type structure, leading to fast access times even 
over very large datasets. Indexing over columns within rows is 
also very powerful, allowing us to quickly retrieve 
rows with particular features.

% Databases but introduce several others. 
One may, of course, load textual data into a relational database, and 
use SQL to retrieve the records of interest. This approach has several 
disadvantages, however. Firstly, one must first examine the data 
construct an appropriate SQL schema to hold the columns 
of interest. It is sometimes feasible to do this automatically 
for data files, but is certainly not a trivial process, 
particularly if the schema 
must be portable across different relational database platforms.
Once the schema has been constructed, the data must be inserted into 
this database line-by-line by constructing SQL code, which is a 
time-consuming and error prone process. Even assuming that the data 
has been inserted into a table with an appropriate schema, there 
are still difficulties to be addressed. Maintaining a database 
server is a difficult task and when the volumes of data become
large, a specialised administrator is usually required to 
ensure the server works correctly.

% We don't need concurrency, etc, as it's always read-only.
Much of the time, a database server is far more complicated than
we require for large scale biological data. Data files 
are usually written once by an application 
and is not subsequently expected to change. Thus, storing this information 
in a relational database with its sophisticated concurrency control
and guarantees of consistency in the face of asynchronous updates
is entirely unnecessary. Even the client-server architecture of 
a typical relational database is wasteful in this case: protocol 
commands must be read and parsed, and responses must be encoded,
written to a socket and subsequently converted back to native types,
all of which constitutes a significant overhead when streaming large 
quantities of data. The limitations of relational database servers 
have been acknowledged for some time now, and has led to a 
variety of `NoSQL' database systems~\citep{m12}.

\section{Wormtable}

Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data with high-performance access to rows via indexes. Tables can
be either written directly via the software library, or converted
from existing formats via command line tools. Tables can be read 
concurrently by any number of processes, and are accessed by a 
simple API in Python.

To take advantage of the data processing capabilities of a database
without the overheads and disadvantages of a relational database 
server, Wormtable is built on top of Berkeley DB~\citep{obs99}.
Berkeley DB is an open source embedded database toolkit, that provides a
scalable key-value store. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}.

Berkeley DB (or simply DB) 
differs from relational database servers in several ways. 
Firstly, it is a software library that a program links against, 
and so there is no need for a separate database server process.
Secondly, DB provides only a key-value store, and so has 
no knowledge of the structure of the records that it stores,
treating them simply as an array of bytes. While there 
are sophisticated 
mechanisms within DB for locking, transactions and so on, 
these are all optional, allowing a very lightweight 
read-only store to be built using the library.

This is the approach taken in Wormtable. We assume that tabular data 
is written once, by some program generating the data, and subsequently 
read many times. Under these assumptions, the primary database 
holding the rows can be opened in a read-only state, allowing any 
number of processes to read it concurrently in a lock free manner.

Since DB is agnostic to the structure of data stored in rows, we 
are free to design our own formats. In Wormtable, the row format is
a compact and flexible binary format holding integer, floating point 
or character values of various sizes. Each column in a row may contain 
either a fixed or variable number of elements (each element being of a
fixed type and size). This row format enables two important properties:
firstly, values from any column can be accessed in constant 
time; and secondly, values in columns can be converted directly into 
native machine values, ready for calculations.

Indexing of columns is achieved via DB also. An index is simply a
database mapping values in a column (or list of columns) to rows 
in the primary database. Indexes allow us to find rows with particular 
properties very quickly, and are a very useful method of excluding 
rows with certain properties from our analyses.

@@

A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
or VCF. In VCF,  information about variant sites in a genome is encoded 
as tab-delimited rows in a text file. To access information about a 
particular site, one must proceed through the file line-by-line until the 
desired record is found. There are methods available to index this 
file~\citep{li11}, but these have no semantic 
understanding of the structure of the file and so are limited in the 
fields that can be indexed. One cannot, for example, build an index 
to quickly obtain all records in which the quality field is greater
than some value. Such queries can take a considerable time on large VCF 
files, which discourages testing and checking for the effects of 
arbitrary thresholds in analyses.

Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples to 
illustrate the improvements over existing methods. To do this, 
we converted a VCF file with XXX rows from the YYY project 
into wormtable format. The (uncompressed) VCF file was XX
gigabytes, and the converted wormtable YY gigabytes, which required 
X hours.

@@

To illustrate the use of indexes in wormtable, 
we created an index on the ALT and REF columns, 
and another on the IN, UT, SY and NS columns. These indexes required 
X and Y MB of disc space and X and Y minutes to build, respectively. 
The ALT+REF index provides a much more efficient means of counting 
the transitions and transversions, and another script to calculate 
these required x seconds. Using the IN+UT+SY+NS index, we 
calculated the mean nucleotide diversity for SNPs (not indels) 
at synonymous, nonsynonymous, intronic and UTR sites. This required 
X seconds. Computing the same values with PyVCF required Y hours.

@@

\section{Conclusions}
Text files are the oldest and most reliable means of exchanging 
information between computers. Plain text, however, does not provide an efficient 
means of storing and processing large volumes of data. Current practice 
in bioinformatics is to encode data in more and more complex plain 
text data formats. While these formats are excellent from the perspective 
of minimising software dependencies, they are very inefficient in terms of 
storage space, the time required to parse them, and the linear
seek times to find particular rows.

Wormtable is not intended to replace text files as the universal 
interchange format. It is intended to provide a persistent data structure 
that efficiently holds data in a form that can be efficiently processed 
and searched. 
% Furthermore, by providing command line programs to import and index common
% formats, we provide a very intuitive and easy to use method to deal
% with standard bioinformatic formats....
Using this data structure, researchers with no knowledge of 
database systems can take full advantage of sophisticated 
data management techniques, and write simple code that processes this 
data in a very high performance manner. In our example above, 
we [blah blah blah]. A systematic application of these techniques 
can substantially increase the researcher's productivity and ability
to explore their data.

Wormtable is an open and collaborative project seeking feedback and 
contributors. The project is in its early stages, currently 
providing only a Python interface. A C library, Perl and Java interfaces
are also planned for Wormtable, which should make the format available
to the majority of bioinformatics projects. With such broad support,
it is hoped that other software tools may use these libraries 
to support direct output of data in Wormtable format.

>>>>

% Databases solve these problems
The classical approach to working with large volumes of 
structured data is to use a database, which solve these 
problems very effectively.
Data is typically encoded in an efficient binary 
format, saving substantial space and avoiding the need
for repeated parsing. Records are usually stored in a
B-tree, leading to fast access times even 
over very large datasets. Indexing over columns within rows is 
also very powerful, allowing us to quickly retrieve 
rows with particular features.

% Databases but introduce several others. 
One may, of course, load textual data into a relational database, and 
use SQL to retrieve the records of interest. This approach has several 
disadvantages, however. Firstly, one must first examine the data 
and construct an appropriate SQL schema to define the appropriate columns. 
It is sometimes feasible to do this automatically 
for data files, but is certainly not a trivial process, 
particularly if the schema 
must be portable across different relational database platforms.
Even assuming that the data 
has been inserted into a table with an appropriate schema, there 
are still difficulties to be addressed. Maintaining a database 
server is a difficult task and when the volumes of data become
large, a specialised administrator is usually required to 
ensure the server works correctly.

% We don't need concurrency, etc, as it's always read-only.
Much of the time, a database server is far more complicated than
we require for biological data. Data files 
are usually written once by an application 
and do not subsequently change. Thus, storing this information 
in a relational database with its sophisticated concurrency control
and guarantees of consistency in the face of asynchronous updates
is entirely unnecessary. Interacting with a database is also a
substantial hurdle, as another lanuage must be learned in order 
to gain access to data. Relational databases, although providing 
powerful data management technologies are far more complex than 
is required, and are therefore little used for the types of 
data that we are interested in.

\section{Wormtable}
% what is wormtable?
Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data of essentially unlimited size. Tables can
be either written directly via the library, or converted
from existing formats via command line tools. Tables can be read 
concurrently by any number of processes, and are accessed by a 
simple API in Python.

% why do we use Berkeley DB?
To take advantage of the data processing capabilities of database
technologies without the overheads and disadvantages of a relational database 
server, wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
Berkeley DB is an open source embedded database toolkit, that provides a
scalable key-value store~\citep{m12}. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}. DB is a highly modular library, providing the 
means to define a very lightweight read-only database, without the 
overheads of a server process

% Wormtable's binary row format 
On this solid foundation of row storage in Berkeley DB, wormtable uses a
compact binary row format designed specifically for biological applications.
Each row contains data from a fixed number of columns, and each column 
holds either a fixed or variable number of elements of a fixed type and size.
Supported types are signed and unsigned integers, 
floating point and character data. To ensure that values are 
stored as compactly as possible, wormtable supports integer values from 
$1$ to $8$ bytes, and the standard single and double precision 
floating point values.

% Wormtable's indexes 
An essential element of database technology is the ability to index tables
to rows with certain properties to be quickly retrieved. Wormtable 
supports indexing over columns or groups of columns within a table, 
making many operations extremely efficient.

@@

% What is VCF?
A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
or VCF. In VCF,  information about variant sites in a genome is encoded 
as tab-delimited rows in a text file. To access information about a 
particular site, one must proceed through the file line-by-line until the 
desired record is found. There are methods available to index this 
file~\citep{li11}, but these have no semantic 
understanding of the structure of the file and so are limited in the 
fields that can be indexed. One cannot, for example, build an index 
to quickly obtain all records in which the quality field is greater
than some value. Such queries can take a considerable time on large VCF 
files, which discourages testing and checking for the effects of 
arbitrary thresholds in analyses.

Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples to 
illustrate the improvements over existing methods. To do this, 
we converted a VCF file with XXX rows from the YYY project 
into wormtable format using the included \texttt{vcf2wt} 
conversion program.  The (uncompressed) VCF file was XX
gigabytes, and the converted wormtable YY gigabytes, which required 
X hours.

@@

To illustrate the use of indexes in wormtable, 
we created an index on the ALT and REF columns, 
This index required X MB of disc space Y minutes to build, respectively. 
The ALT+REF index provides a much more efficient means of counting 
the transitions and transversions, and another script to calculate 
these required x seconds.

@@

\section{Conclusions}
% text files are good for archiving, but terrible for processing
Text files are the oldest and most reliable means of exchanging 
information between computers. Plain text, however, does not provide an efficient 
means of storing and processing large volumes of data. Current practice 
in bioinformatics is to encode data in more and more complex plain 
text data formats. While these formats are excellent from the perspective 
of minimising software dependencies, they are very inefficient in terms of 
storage space, the time required to parse them, and the linear
seek times to find particular rows.

% We're not trying to define a universal bioinformatics file format,
% just to help people be more productive.
Wormtable is not intended to replace text files as the universal 
interchange format. It is intended to provide a persistent data structure 
that efficiently holds data in a form that can be efficiently processed 
and searched. Using this data structure, researchers with no knowledge of 
database systems can take full advantage of sophisticated 
data management techniques, and write simple code that processes this 
data in a very high performance manner. 
By providing conversion programs to convert 
common formats to wormtable, we also provide an easy to 
use common platform for 
bioinformatics analysis. This reduces the need for third party libraries 
to parse complex files and considerably simplifies the code required 
to process data.
Together, these advantages of increased performance and reduced 
code complexity can substantially increase a researcher's 
productivity and ability to explore their data.

% If we get some help, it could be pretty awesome...
Wormtable is an open and collaborative project seeking feedback and 
contributors. The project is in its early stages, currently 
providing a conversion tool for the VCF format only. With community
participation it may grow to support many other formats and become
a powerful common platform for bioinformatics data analysis.
